---
title: "Portfolio Development and Planning"
author: "Kate OBrien"
format: html
echo: false
message: false
warning: false
results: false
---

## Dataset

https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-10-08/readme.md#most_visited_nps_species_datacsv

```{r}
nps_species_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-08/most_visited_nps_species_data.csv')
```

## Data Description

My chosen dataset describes species in the most visited national parks in the US.

Variables and Types: ParkCode- character ParkName- character CategoryName- character Order- character Family- character TaxonRecordStatus- character SciName- character CommonNames- character Synonyms- character ParkAccepted- logical Sensitive- logical RecordStatus- character Occurrence- character OccurenceTags- character Nativeness- character NativenessTags- character Abundance- character NPSTags- character ParkTags- character References- doouble Observations- double Vouchers- double ExternalLinks- character TEStatus- character StateStatus- character OzoneSensitiveStatus- character GRank- character SRank- character

The research question I want to answer is "What is the relationship between sensitive native species and invasive non-native species?

## Data Visualizations

For my visualizations, I'd like to have a comparative scatter chart for a few select national parks to visualize the number of sensitive native species compared to the number of invasive, non-native species. I would also like to create a visualization that demonstrates potential changes in concentrations of native species in each NP overtime. I may need to find additional data about species sightings in national parks by year.



```{r}
library(tidyverse)
```

## Data Cleaning

Here, I made sure to load in all of my observations datasets from Olympic, Joshua Tree, and Yosemite National Parks. I was a little bit careless in installation and had a couple extra variables in some of the sets that I didn't need, and in order to join the rows onto each other, I needed each dataset to have the same variables. I made sure to select the variables I wanted to use so that my observations sets were clean.

After that, I needed to make sure each observations data set had a variable that indicated which park is observed in, so I mutated the data to add a variable with where the observation was made. Once I had that variable, I joined all the datasets together and made one giant observations set.

```{r}
olym_np_obs <- read_csv("data/olym_np_observations.csv")
joshtree_np_obs <- read_csv("data/joshua_tree_np_observations.csv")
yosemite_np_obs <- read_csv("data/yosemite_np_obs.csv")
```

```{r}
#| label: Cleaning NPS Species Data
#nps_species_data |>
#count(ParkName) #helping me figure out which parks I wanna focus on


#nps_species_clean <- nps_species_data |>
 # select(c(ParkName, 
  #         SciName, 
   #        Nativeness, 
    #       CategoryName, 
     #      Observations)) |>
  #filter(CategoryName == "Bird") |> #choosing to focus on birds for this project; going to join a dataset from iNaturalist about observations in specific NPs
  #filter(ParkName == c("Olympic National Park", "Yosemite National Park", "Joshua Tree National Park"))  #Chose 3 NPs to focus on
#write_csv(nps_species_clean, file = "nps_species_clean.csv")
nps_species_clean <- read_csv("data/nps_species_clean.csv")

```

# Building Observations Dataset

```{r}
#| label: Ensuring Obs Datasets Match
olym_np_obs <- olym_np_obs |>
  select(-time_observed_at)

#yosemite_np_obs <- yosemite_np_obs |>
  #select(-taxon_id)
olym_np_obs <- olym_np_obs |>
  mutate(observed_in = "Olympic National Park") |>
  #select(-time_observed_at) |>
  select(-iconic_taxon_name)

joshtree_np_obs <- joshtree_np_obs |>
  mutate(observed_in = "Joshua Tree National Park") |>
  select(-time_observed_at)

yosemite_np_obs <- yosemite_np_obs |>
  mutate(observed_in = "Yosemite National Park") |>
  select(-user_login)
#All observation data taken from iNaturalist!
```



```{r}
np_observations <- rows_insert(joshtree_np_obs, 
                               olym_np_obs)
np_observations <- rows_insert(np_observations, 
                               yosemite_np_obs)
write_csv(np_observations, file = "np_observations.csv")
```



```{r}
np_obs <- read_csv('np_observations.csv')

np_obs_year <- np_obs |>
  mutate(observed_on = ymd(observed_on)) |>
  mutate(year = year(observed_on))
```


```{r}
nps_species_clean <- read_csv("data/nps_species_clean.csv")

nps_species_obs <- inner_join(np_obs_year, nps_species_clean, join_by(scientific_name == SciName))
```

```{r}
#| label: Get rid of unnecessary variables
nps_species_clean <- nps_species_clean |>
  select(-CategoryName)

nps_species_clean %>% 
  filter(str_detect(ParkName, "Yosemite")) %>% 
  count(SciName)

np_obs_year %>% 
   filter(str_detect(observed_in, "Yosemite")) %>% 
  count(scientific_name)

np_species_obs <- np_obs_year |>
  left_join(nps_species_clean, 
            by = c("scientific_name" = "SciName",
                   "observed_in" = "ParkName"))

```
# Building Visuals

## Visual #1

This visual is a comparison of the observations made of Native vs. Non native species in each park. My hypothesis was that as the number of non-native species goes up, the number of native species observations would go down. This may have been a more effective test on plants, or something similar, however, as birds are highly migratory and may not be classified as native or non-native.

```{r}
np_species_obs |>
  group_by(Nativeness, 
           year, 
           observed_in) |>
  summarise(total_observations = 
              sum(Observations, 
                  na.rm = TRUE)) |>
  ggplot(mapping = aes(x = year, 
                       y = total_observations,
                       fill = Nativeness)) +
  geom_col() +
  facet_wrap(~observed_in, ncol = 1, scale = "free") +
  labs(title = "Observations of Native vs. Non-Native Bird Species in Olympic, Yosemite, and Joshua Tree National Park") +
  xlab("Year Observed") +
  ylab("Number of Observations")
ggsave("bar_plot_obs.png", device = png)
```

## Visual #2

After noticing how many 'unknowns' there were in my Nativeness variable, I wanted to make a visual illustrating where the NA values were in the data, maybe to gain a better understanding of what might've happened in data collection. Maybe in recent years, there have been more sightings of various species in the parks with changing climates, but further analysis would be required to determine.

```{r}
np_species_obs |>
  group_by(year,
           observed_in,
           Nativeness) |>
  summarise(missing = 
              sum(is.na(Nativeness))) |>
  ggplot(mapping = aes(x = year,
                       y = missing,
                       colour = observed_in,
                       shape = observed_in)) +
  geom_point(alpha = 0.7) +
  labs(title = "Frequency of Missing Nativeness Values in NP Data",
       x = "Year",
       y = "Number of Missing Values",
       colour = "Park Data") +
  scale_color_viridis_d()
```
